The interactions of rational pragmatic agents provide a framework for understanding efficient language structure and use

### 2019-01-03 paper/paper.Rmd
updates to intro -- now much more focus on Zipf, probably need to carry this throughout.

# Introduction
Why do natural languages look they way they do? Functionalist approaches to language science bring a utility-based perspective to this question, typically reframing it in terms of information-theoretic principles of efficiency -- how does having property $x$ make using language $\ell$ more or less useful for communication? This family of theories has described a range of phenomena in language-design (compositionality, ambiguity, dependency length) and language-use (redundancy and reduction) in terms of communicative efficiency. Consider two examples of such theories: Piantadosi et al. (2012) argued that efficiently *designed* linguistic systems will contain ambiguous lexical items when communication is contextualized. If words with multiple meanings can be disambiguated in context, less costly forms can be re-used efficiently.  From the language *use* perspective, Levy & Jaeger (2007) showed that speakers are sensitive to spikes and troughs in the information rate of their utterances. The authors showed that speakers modulate use of optional material according to the surprisal of oncoming material -- an optimal strategy for efficient information transfer from an information-theoretic analysis.\par

While Piantadosi et al. (2012) and Levy & Jaeger (2007) provide two recent examples of functionalist theory, these frameworks are largely in line with earlier work by Zipf (1949). Zipf's Principle of Least Effort is a domain-general framework for describing human behavior in terms of effort minimization. Zipf claimed that humans have a preference to minimize "the average rate of work-expenditure over time" (Zipf, 1949, p.6). In terms of language behavior, he described a relation between a word's frequency in corpora and its length (number of characters) in terms of the principle of least effort. Intuitively, if effort is related to a word’s length (perhaps via articulatory effort) then, holding all else constant, a language with shorter words should be more efficient than a language with longer words, what Zipf called the *Law of Abbreviation*. Indeed this relation between frequency and length is well-documented in natural languages (Kanwal et al 2017).\par

Perhaps even more important than the actual predictions made by Zipf’s Principle of Least Effort is the insight that we can analyse the structure of language systems in terms *effort* or *cost* along competing dimensions.  With such a formulation we can begin to ask questions about what optimal languages might look like under different cost structures. For example, Zipf noted that purely from the standpoint of speaker effort, what Zipf called "Speaker's Economy," an optimal language $\ell_{speaker}^*$ would tend toward a vocabulary of just a single, low-cost word. Given such a language, the full set of potential meanings would be conveyed using only that word, i.e. $\ell_{speaker}^*$ would be fully ambiguous and all possible meanings would need to be disambiguated by a listener. Conversely, from the standpoint of listener effort, what Zipf called "Auditor's Economy," an optimal language $\ell_{listener}^*$ would bijectively map all possible meanings to distinct words, eliminating a listener's need to disambiguate. Under this analysis, speaker effort is related to production cost and listener effort to understanding (disambiguation) cost. Clearly natural languages fall between the two extremes of $\ell_{speaker}^*$ and $\ell_{listener}^*$. Zipf's insight was that the particular distributional properties we observe in natural languages emerge from the competition of these forces -- the pressure to jointly minimize speaker and listener effort.\par

Zipf’s principle of least effort and its speaker and listener instantations were intended as a framework for explaining all of natural language. However, Zipf’s primary unit of analysis was at the level of languages themselves (i.e. the vocabulary). Subsequent work by Horn (1984), highlighted the importance of Zipf’s principles for explaining phenomena at the level of a single conversation. Horn suggested a direct link between Zipf’s Speaker and Listener economy and, what was at the time, more recent work on conversational pragmatics by Grice (1975). Horn highlighted that the interaction of Zipf’s forces were “largely responsible for generating Grice’s conversational maxims and the schema for pragmatic inference derived therefrom.” Put differently, system-level efficiency we see in languages is deeply related to local-level efficiency during the in-the-moment interactions of speakers and listeners. \par

In this work we present an initial step toward formalizing this connection first highlighted by Horn. That is, we present a framework for evaluating the aggregate, distributional properties of linguistic systems as emerging from the accumulation of many in-the-moment interactions of pragmatic speaker and listener agents. We are exploring the questions - to what extent can we connect the Gricean and Zipfian programs in a single framework? To do so, we adopt a simulation-based approach, modeling communication as reference games in which we can vary the types of agents and languages present. This framework allows us to investigate a range of functionalist theories broadly under the categories of efficient language design and language use. \par

For this particular project we focus on an analysis of the communicative function of ambiguity in language following work by Piantadosi et al. (2012). We adopt this particular linguistic phenomena to demonstrate the extent to which we can explore both the design- and use- based functionalist ideas. We focus on three questions -- (1) when is it desirable to have a language with lexical ambiguity? (2) Under what circumstances do speakers and listeners use ambiguity efficiently? (3) Assuming Gricean agents what is the impact of increasing pragmatics on efficiency?\par

We begin with a high-level introduction to the modelling framework introducing the basic ingredients we will need to represent language as repeated reference games. Following this introduction we derive a simple objective function for measuring the efficiency of linguistics systems in this setting. Subsequently, we move on to three case-studies examining the questions posed above, framing results in terms of our efficiency measure.

@@

To assess the optimality of a linguistic system we need some measure of goodness. In other words an objective function, which, as a function of the particular linguistic system $\ell$ and the speaker $S$ and listener $L$ agents returns a measure of the system efficiency ( $f(S, L, \ell) \rightarrow \mathbb{R}$).  For our purposes we’d like to tie this objective to fundamental principles of speaker and listener effort suggested by Zipf. Additionally, we will take the fundamental unit of analysis as the reference game event. \par
We now derive a simple objective function from Zipfian first principles. Zipf proposed that the particular distributional properties found in natural language emerge as a result of competing speaker and listener pressures. We operationalize this in equation (1) -- the efficiency of a linguistic system $\ell$ being used by speaker and listener agents $S$ and $L$ is sum of the expected speaker and listener effort to communicate.
\begin{equation}
\text{Efficiency}(\ell) = \mathbb{E}_{e \in E}[\text{speaker effort}] + \\ \mathbb{E}_{e \in E}[\text{listener effort}]
\end{equation}

@@

Rewriting (1) we now have
\begin{equation}
= \mathbb{E}_{e \in E}[-log_2(p(u))] + \\
\mathbb{E}_{e \in E}[-log_2(L(m|u; \ell))]
\end{equation}
In general, these expectations are each taken over all possible communicative events $e\in E$ weighted by the probability of a particular event $p(e)$. Recall that is the set of all utterance, meaning pairs $<u, m> = e \in E$.

@@

# Three case-studies on efficient language use and design
We now have a metric for exploring questions of langauge optimality in the reference game setting (the speaker-listener cross-entropy). In the following sections we examine the behavior of the objective with respect to the communicative function of ambiguity. In particular, we examine three questions. The first is a question of optimal design -- given a set of possible languages $L$ in what circumstances does the optimal languages $\ell^*$ contain ambiguity? The second is a question of optimal *use* -- give a particular language that theoretically allows for the use of ambiguious material under what circumstances will ambiguous forms be used efficiently?

@@

Piantadosi et al. (2012) provided empirical evidence for this claim, showing (SAY WHAT PIANTADOSI DID). We consider their basic arguemnt in the reference game setting. Our approach procceds as follows, we consider the full-space of valid languages (semantic mappings) between forms and utterances in which $|U|=|M|=4$. A language $\ell \in L$ is "valid" so long as each possible meaning in $m \in M$ can be referred to by at least one form $u \in U$ (every column of $\ell$ has some non-zero assigment). For example, a one word language that is fully ambiguous would map a single form to all possible meanings. We represent that via the boolean matrix $\ell_1$ in which $[[u_1]] = \{m_1,m_2, m_3 \}, [[u_2]]=[[u_3]]=\{\}$

@@

### Results
We run $n=100$ simulations in four condtions. Our first is a *single-context* condition ( $|C|=1$) -- there is a only a single context describing $p(m|c_1)$. Our second condition contains two-contexts  ($|C| = 2$) -- we consider efficiency under both $p(m|c_1)$ as well as $p(m|c_2)$. The third and fourth condition correspond accordingly with $|C|=3$ and $|C| = 4$, respectively. In each condition we consider a random set of utterances costs $P(u)$ and need probabilities $p(m|c)$ parametrized as Dirichlet distributions with $\alpha=1$. Piantadosi et al. (2012) argued that ambiguity is an efficient property of contextualized langauge use. That is, it is useful to assign words multiple meanings if those words can be disambiguated in context. Formalized in our setting, this would mean that we expect $\ell*$ to be more likely to contain ambiguous mappings as the number of contexts increases.par
Figure 1. plots the proportion of optimal languages under $H_{cross}(P_{Speaker}, P_{Listener})$ for each condition. We find that as the number of contexts increases, so does the probability that optimal language $\ell*$ contains ambiguity. For comparison we also include the optimal language under an objective that only considers speaker or listener effort. In line, with Zipf's predictions, if langauges are designed only to minimize speaker effort then optimal languages will assign all meanings to a single, low-cost utterance. Likewise, if langauges are designed only to minimize speaker listener effor then ambiguity should alwasy be avoided.

>>>>

# Introduction

Why do natural languages look they way they do? While Zipf (1935) presented his “Principle of Least effort” as a domain general framework for understanding human behavior, he took language as his central case-study. He proposed that distributional properties found in natural language were evidence of speaker-listener utility maximization or, equivalently, effort minimization. In his own words, “we are arguing that people do in fact act with a maximum economy of effort, and that therefore in the process of speaking-listening they will automatically minimize the expenditure of effort.” Evidence supporting this hypothesis has been realized in a set of well-known findings, largely falling under what is now known as Zipf’s Law or Zipfian distributions -- one of a family of discrete power law distributions. A range of empirical phenomena in social and physical sciences are well-described as Zipfian, from the relationship between word rank and frequency (CITATION) to the populations of U.S. cities (CITATION) Stanford stats class 306a).\par

Language-structure and use provide a particularly fascinating case-study under an efficiency-based analysis as speaker- and listener-effort is asymmetric. Simply put, in the act of communicating with another person, what is effortful as a speaker is likely different from what is effortful as a listener. For example, Zipf noted that purely from the standpoint of speaker effort, what Zipf called "Speaker's Economy," an optimal language $\ell_{speaker}^*$ would tend toward a vocabulary of just a single, low-cost word. Given such a language, the full set of potential meanings would be conveyed using only that word, i.e. $\ell_{speaker}^*$ would be fully ambiguous and all possible meanings would need to be disambiguated by a listener. Conversely, from the standpoint of listener effort, what Zipf called "Auditor's Economy," an optimal language $\ell_{listener}^*$ would bijectively map all possible meanings to distinct words, eliminating a listener's need to disambiguate. Under this analysis, speaker effort is related to production cost and listener effort to understanding (disambiguation) cost. Clearly natural languages fall between the two extremes of $\ell_{speaker}^*$ and $\ell_{listener}^*$. Zipf’s Law of Meaning proposes that the particular relation between a word’s frequency and denotation size emerge from the competition of these forces -- the pressure to jointly minimize speaker and listener effort.\par

A range of more recent approaches to language science have distinctly Zipfian roots (functionalist theories). In terms of our opening question, “why do languages look the way they do?”, a functionalist approach would formalize the question in terms of information-theoretic principles of efficiency -- how does having property $x$ make using language $\ell$ more or less useful for communication? This family of theories has described a range of phenomena in language-design (compositionality, ambiguity, dependency length) and language-use (redundancy and reduction) in terms of communicative efficiency. Of particular interest to our current work, Piantadosi et al. (2012) argued that efficiently *designed* linguistic systems will contain ambiguous lexical items when communication is contextualized. If words with multiple meanings can be disambiguated in context, less costly forms can be re-used efficiently. They showed in a series of analyses that … [WHAT PIANTADOSI DID HERE!!!!]. \par

Zipf’s principle of least effort and it’s associated speaker and listener instantations were intended as a framework for explaining all of natural language. Evidence for the theory was largely derived from vocabulary-level analysis. Implicit in Zipf’s (1935) argument is that aggregate, distributional-level effects emerge from optimization during local interactions between speaker-listeners. While Functionalist projects have provided evidence that speaker’s attempt to optimize their output at a variety of levels -- phonologically (morphological contraction; Bell et al., 2003), syntactically (Levy & Jaeger, 2007; Jaeger, 2010) and over even over discourse (Genzel & Charniak, 2002), Zipf was less precise in terms of his exact level of analysis -- however his example of optimal speaker and listener language in which a single word $u$ can express some meaning $m$$[[u]]\neq \emptyset$.

A half century after Zipf, the linguist Lawrence Horn (1984), highlighted the importance of Zipf’s principles for explaining conversation-level phenomena. Horn suggested a direct link between Zipf’s Speaker and Listener economy and, what was at the time, more recent work on conversational pragmatics by Grice (1975). Horn highlighted that the interaction of Zipf’s forces were “largely responsible for generating Grice’s conversational maxims and the schema for pragmatic inference derived therefrom.” Put differently, system-level efficiency we see in languages is deeply related to local-level efficiency during the in-the-moment interactions of speakers and listeners. \par

In this work we present a step toward formalizing this connection highlighted by Horn -- formalizing the theoretical connection between Zipfian and Grice, and in doing so, deriving aggregate, distributional level properties from local, in-the-moment interactions of speaker-listeners. To gain traction on this project we adopt a simulation-based approach, modeling communication as reference games in which we can vary the types of agents and languages present. This framework allows us to investigate a range of functionalist theories broadly under the categories of efficient language design and language use. \par

For this particular project we focus on an analysis of the communicative function of ambiguity in language following work by Piantadosi et al. (2012). We adopt this particular linguistic phenomena to demonstrate the extent to which we can explore both the design- and use- based functionalist ideas. We focus on three questions -- (1) when is it desirable to have a language with lexical ambiguity? (2) Under what circumstances do speakers and listeners use ambiguity efficiently?\par

We begin with a high-level introduction to the modelling framework introducing the basic ingredients we will need to represent language as repeated reference games. Following this introduction we derive a simple objective function for measuring the efficiency of linguistics systems in this setting. Subsequently, we move on to two case-studies examining the questions posed above, framing results in terms of our efficiency measure.

@@

To assess the optimality of a linguistic system we need some measure of goodness. In other words an objective function, which, as a function of the particular linguistic system $\ell$ and the speaker $S$ and listener $L$ agents returns a measure of the system efficiency ( $f(S, L, \ell) \rightarrow \mathbb{R}$).  For our purposes we’d like to tie this objective to fundamental principles of speaker and listener effort suggested by Zipf. Additionally, we will take the fundamental unit of analysis as the reference game event. \par

We now derive an objective function from Zipfian first principles. Zipf proposed that the particular distributional properties found in natural language emerge as a result of competing speaker and listener pressures. We operationalize this in equation (1) -- the efficiency of a linguistic system $\ell$ being used by speaker and listener agents $S$ and $L$ is sum of the expected speaker and listener effort to communicate over all possible communicative events.

\begin{equation}
\text{Efficiency}(S, L, \ell) = \mathbb{E}_{e \in E}[\text{speaker effort}] + \\ \mathbb{E}_{e \in E}[\text{listener effort}]
\end{equation}

@@

Rewriting (1) we now have
\begin{equation}
\begin{split}
\text{Efficiency}(S, L, \ell) = \mathbb{E}_{e \in E}[-log_2(p(u))] + \\
\mathbb{E}_{e \in E}[-log_2(L(m|u; \ell))]
\end{split}
\end{equation}
In general, these expectations are each taken over all possible communicative events $e\in E$ weighted by the probability of a particular event $p(e)$. Recall that is the set of all utterance, meaning pairs $<u, m> = e \in E$.

@@

# Two case-studies on efficient language use and design
We now have a metric for exploring questions of language optimality in the reference game setting (the speaker-listener cross-entropy). In the following sections we examine the behavior of the objective with respect to the communicative function of ambiguity. In particular, we examine three questions. The first is a question of optimal design -- given a set of possible languages $L$ in what circumstances does the optimal languages $\ell^*$ contain ambiguity? The second is a question of optimal *use* -- give a particular language that theoretically allows for the use of ambiguous material under what circumstances will ambiguous forms be used efficiently?

@@

Piantadosi et al. (2012) provided empirical evidence for this claim, showing (SAY WHAT PIANTADOSI DID). We consider their basic argument in the reference game setting examining the full-space of valid (footnote here for “valid” we have the constraint that all meanings must be expressible) languages (semantic mappings) between forms and utterances in which $|U|=|M|=4$. A language $\ell \in L$ is "valid" so long as each possible meaning in $m \in M$ can be referred to by at least one form $u \in U$ (every column of $\ell$ has some non-zero assignment). For example, a one word language that is fully ambiguous would map a single form to all possible meanings. We represent that via the boolean matrix $\ell_1$ in which $[[u_1]] = \{m_1,m_2, m_3 \}, [[u_2]]=[[u_3]]=\{\}$

@@

### Results
We run $n=100$ simulations in four condtions. Our first is a *single-context* condition ( $|C|=1$) -- there is a only a single context describing $p(m|c_1)$. Our second condition contains two-contexts  ($|C| = 2$) -- we consider efficiency under both $p(m|c_1)$ as well as $p(m|c_2)$. The third and fourth condition correspond accordingly with $|C|=3$ and $|C| = 4$, respectively. In each condition we consider a random set of utterances costs $P(u)$ and need probabilities $p(m|c)$ parametrized as Dirichlet distributions with $\alpha=1$. Piantadosi et al. (2012) argued that ambiguity is an efficient property of contextualized langauge use. That is, it is useful to assign words multiple meanings if those words can be disambiguated in context. Formalized in our setting, this would mean that we expect $\ell*$ to be more likely to contain ambiguous mappings as the number of contexts increases.\par
Figure 1. plots the proportion of optimal languages under $H_{cross}(P_{Speaker}, P_{Listener})$ for each condition. We find that as the number of contexts increases, so does the probability that optimal language $\ell*$ contains ambiguity. For comparison we also include the optimal language under an objective that only considers speaker or listener effort. In line, with Zipf's predictions, if langauges are designed only to minimize speaker effort then optimal languages will assign all meanings to a single, low-cost utterance. Likewise, if langauges are designed only to minimize speaker listener effor then ambiguity should alwasy be avoided.
